---
title: "BikeShare Capstone"
author: "Olaposi"
date: "2023-12-29"
output: 
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
chooseCRANmirror(graphics=FALSE, ind=1)
knitr::opts_chunk$set(echo = TRUE)
```

## Bike Share Capstone

This is an analysis to complete the Google Data Analystics course. The purpose of this analyis is to answer the question: "How do annual members and casual riders use Cyclistic bikes differently?"


### Install/Load Needed Packages & Importing Data in R

In order to run this analysis in R, a few packages are needed.

```{r installing packages, include=FALSE}
install.packages("tidyverse")
library(tidyverse)
installed.packages("lubridate")
library(lubridate)
install.packages("ggplot2")
library(ggplot2)
install.packages("readr")
library(readr)
install.packages("dplyr")
library(dplyr)
```


After installing these packages, I downloaded the data I would be working with. [Bike Share Data](https://divvy-tripdata.s3.amazonaws.com/index.html) I will be using the q2_2019, q3_2019, q4_2019, & q1_2020 datasets. 


After downloading these datasets, I created a folder to store them in. Once create I needed to check and see if I was working in the right file location. It was not, so adjusted it accordingingly. 

```{r Working directory, include=FALSE}
getwd() #this checks where you are
setwd("C:/Users/Lyrical Hero/Documents/Bike_Capstone") #this changes to desired directory
```


```{r Importing data, include=FALSE}
q2_2019 <- read_csv("Divvy_Trips_2019_Q2.csv")
q3_2019 <- read_csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("Divvy_Trips_2019_Q4.csv")
q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")
```


### Tranforming the Data 

Once the data is pulled, its best practice to inspect it. I checked the column names of each dataframe and compared them. Below I see the column names are all over the place. I plan to merge these datasets so the names of the columns need to be the same. This will give me one complete and concise file.

```{r}
colnames(q2_2019)
colnames(q3_2019)
colnames(q4_2019)
colnames(q1_2020)
```

After checking the column names, it's clear to see that there are some changes that need to be made. I then renamed the columns in each table to match the columns in the q1_2020. This is due to q1_2020 being the most recent data.

```{r renaming the columns, include=FALSE}
q4_2019 <- rename(q4_2019, ride_id=trip_id,rideable_type = bikeid,
                  started_at=start_time, ended_at=end_time, start_station_name=from_station_name,
                  start_station_id=from_station_id,end_station_name = to_station_name, 
                  end_station_id = to_station_id,member_casual = usertype)


q3_2019 <- rename(q3_2019 ,ride_id = trip_id, rideable_type = bikeid, started_at = start_time,  
                  ended_at = end_time, start_station_name = from_station_name, 
                  start_station_id = from_station_id, end_station_name = to_station_name, 
                  ,end_station_id = to_station_id, member_casual = usertype)

q2_2019 <- rename(q2_2019, ride_id = "01 - Rental Details Rental ID",
                  rideable_type = "01 - Rental Details Bike ID",
                  started_at = "01 - Rental Details Local Start Time",  
                  ended_at = "01 - Rental Details Local End Time",  
                  start_station_name = "03 - Rental Start Station Name", 
                  start_station_id = "03 - Rental Start Station ID",
                  end_station_name = "02 - Rental End Station Name", 
                  end_station_id = "02 - Rental End Station ID",
                  member_casual = "User Type")
```


Up next I examined the dataframes to identify any inconsistencies.

```{r}
str(q1_2020)
str(q4_2019)
str(q3_2019)
str(q2_2019)
```

Here I saw that ride_id and ride_type are not consistent throughout each of the dataframes. So I used q1_2020 as the standard and change the column's data types for the other three dataframes.

```{r, include=FALSE}
q4_2019 <-  mutate(q4_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
q3_2019 <-  mutate(q3_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
q2_2019 <-  mutate(q2_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
```


From there I combined all of the dataframes into one easy to work with dataframe.

```{r merging dataframes, include=FALSE}
all_trips <- bind_rows(q2_2019, q3_2019, q4_2019, q1_2020)
```

Lastly, I removed all sensitive and identifiable data. This is in order to protect the riders privacy.

```{r removing unneeded columns, include=FALSE}
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, "01 - Rental Details Duration In Seconds Uncapped", "05 - Member Details Member Birthday Year", "Member Gender", "tripduration"))

```

### Cleaning the Data

After looking through the table, there were a few things that needed to be cleaned.<br/>
1. The member_casual have 4 different options rather than the 2 given in q1_2020.<br/>
2. The start and end gives all the data in a format that isn't useful.<br/>
3. The q1_2020 did not have trip duration, so we have to include one for the all_trip.<br/> 
4. We will need to delete any of the trips that show negative rides.<br/>

#### 1. Membership Inconsistencies

Let's start with checking the tables for member_casual, There are 4 option: casual, Customer, member, Subscriber. 

```{r}
table(all_trips$member_casual)
```

To keep things inline with the current format, I reassigned the older values to be inline with the q1_2020 column.

```{r}
all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual
                                ,"Subscriber" = "member"
                                ,"Customer" = "casual"))
```

Check the member_casual column once again to make sure that there is now only "member" and "casual".

```{r}
table(all_trips$member_casual)
```

#### 2. Fixing the date and time

The started_at and ended_at columns has so much infomation merged together. With this information in this state, it would be very hard to work with this data. Therefore, I created a date column. From there I split the date into different columns to make date easy to use. It was split into, month, day, year, and the day of the week.

```{r}
all_trips$date <- as.Date(all_trips$started_at) 
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

#### 3. Trip Duration Problem

There is nothing to show how long each ride was directly. However we do have the start and end times of each ride so can use that, to figure out each trip's duration. I created a column that calculated the time of each ride. This column is called ride_length.

```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```

However, due to binding 4 table together, the data types did not all match. So I converted the ride_length to be a numeric type across the entire dataframe.

```{r include=FALSE}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.character(all_trips$ride_length)
all_trips$ride_length <- as.numeric(all_trips$ride_length)
is.numeric(all_trips$ride_length)
```

#### 4. Removing Bad Data

Finally for cleaning I removed the negative ride times, as well removing the bikes that were taken out of circulation.

```{r}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```



### Analysis


Time to run a summary on the ride_length. 

```{r}
summary(all_trips_v2$ride_length)
```

I also wanted to compare members and casual user according to thier ride lengths.
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```


After checking checking the ride time for the user type and and, I noticed the days were out of order.  So I ordered the days and then checked again.

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```


After, it was time to check the average ride time and number of rides. This is split up by day of the week. and membership type.

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n() ,average_duration = mean(ride_length)) %>%
   arrange(member_casual, weekday)	
```

From here, it's now time to add visuals. 

```{r}
all_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(),
            average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Number of Rides by Day of the Week and Member Type",
       x = "Day of the Week",
       y = "Number of Rides") +
  theme_minimal()
```


```{r}
all_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(),
            average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday)  %>%
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Average Ride Duration by Day of the Week and Member Type",
       x = "Day of the Week",
       y = "Average Ride Duration (sec)") +
  theme_minimal()
```

### Conclusion 

After running the analysis I was able to come to some data driven conclusion to the question, "How do annual members and casual riders use Cyclistic bikes differently?" 

1. Members tend to ride use the bike sharing service much more frequently than casual users each day of the week. 

2. Casual users on average tend to use the bike sharing service for longer periods of time *(on average over 3x longer)* compared to members. 

3. On week days, members rides are about 4x the amount of casual user. And on the weekend those numbers start to even out. 

 
